---
title: "ALSPAC.helpR_tutorial"
author: "Serena Defina"
date: "`r Sys.Date()`"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 90
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Hey,

This tiny tutorial is to help you get started with
[**`ALSPAC.helpR`**](https://github.com/SereDef/ALSPAC.helpR). The package collects a few
small functions to load, clean, and manipulate datasets that may come in handy when
working with ALSPAC data.

## Installation

You can install the development version of **`ALSPAC.helpR`** from
[GitHub](https://github.com/) with:

```{r}
# install.packages("devtools")
devtools::install_github("SereDef/ALSPAC.helpR")

# Lets load it up 
library(ALSPAC.helpR)
```

## 1. Load the data

First thing you usually need to do is *read* the dataset in memory. So meet our first
function: **`load_alspac()`** that can do a few useful things: - reads .sav format - about
15.500 rows and maybe also loads of columns (takes a second)

```{r, eval=F, echo=T}
?load_alspac

data <- load_alspac() # Opens interactive window to choose file
# Note: by default, the function transforms all variable names to lowercase (set lower.case=FALSE if this is not desired)
```

## Explore dataset

```{r, eval=F, echo=T}
# Find all variables in "data" whose names start with "f01"
find_var('f01',  method='starts')
# OR, to save some typing 
fv('f01',  method='starts')

# Find all the variable labels that mention "income"
find_lab('income')
# OR, to save some typing 
fl('income')

# Seelect some variables
select_var('depre', times=c(1,3)) # returns, for example c('depre_1','depre_3')
# OR, to save some typing 
sel('depre')

```

## Clean dataset

```{r, eval=F, echo=T}
# Create unique identifier
data$IDC <- make_idc(mom.id='id123') 
# Create sex factor
data$sex <- make_sex_factor()

# Remove one of the sibling couple randomly
data_nosibl <- rm_siblings(method='random')

# Remove outliers 
data$horses <- rm_outliers('unicorns', cutoff=3.5)
```
