---
title: "ALSPAC.helpR_tutorial"
author: "Serena Defina"
date: "`r Sys.Date()`"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 90
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Hey,

This tiny tutorial is to help you get started with
[**`ALSPAC.helpR`**](https://github.com/SereDef/ALSPAC.helpR). The package collects a few
small functions to load, clean, and manipulate datasets that may come in handy when
working with ALSPAC data.

## Installation

You can install the development version of **`ALSPAC.helpR`** from
[GitHub](https://github.com/) with:

```{r}
# install.packages("devtools")
devtools::install_github("SereDef/ALSPAC.helpR")

# Lets load it up 
library(ALSPAC.helpR)
```

## 1. Load the data

First thing you usually need to do is *read* the dataset in memory. So meet our first
function: **`load_alspac()`** that can do a few useful things:

-   reads SPSS (`.sav`) format
-   typically, almost 16.000 rows and maybe also loads of columns (\~10.000), so give him
    a minute, will tell you when he is done

[Defaults]{.underline}:

-   opens interactive window unless `filepath` is specified
-   `lower.case = TRUE` all column names are lower cased (also the metadata)
-   `keep.value.labels = TRUE` let's talk about the format of ALSPAC data (and SPSS)
-   `load.metadata = FALSE` metadata is lazy by default

[Tip]{.underline}: call the dataset "data", see why in a minute...

```{r loading}
# ?load_alspac

data <- load_alspac( # filepath = NULL, 
  lower.case = TRUE,
  keep.value.labels = TRUE,
  load.metadata = FALSE
)
```

#### The metadata file

This is just a dataframe taken from the ALSPAC website (aka what you see on the variable
search tool) Why having this in your R session?

-   this is filtered for the variables that you have in the dataset
-   it has additional information about age of measurement
-   you can use it to inspect the data or embed metadata information in your script

```{r}
?alspac_metadata

# No data file available, just want to inspect _all_ data?
# No problem 
all_metadata <- ALSPAC.helpR::alspac_metadata
```

## 2. Explore dataset

OK, we've got the data, let's do something fun with it now.

Whether you load metadata explicitly or not, you can use it to explore the dataset using
two functions: **`find_var()`** and **`find_lab()`**

-   They each have a shortcut because I fully endorse lazy typing: `fv()` and `fl()`
-   Also if you named your dataset "data" (like I told you) you win extra non-typing time
    (use the `data` argument otherwise)
-   They take a string, and understand *regex* (but can also use the `method` argument if
    you don't remember, possible values: "starts", "ends", "contains")
-   `print.labels = FALSE` will return a vector of variable names (that you can use for
    example for transforming) rather than the dataframe

```{r}
# ?find_var

# Find all variables names in "data" whose names start with "c2"
find_var('c2', 
         # data = data,
         method='starts', 
         print.labels = TRUE,
         to.data.frame = FALSE)
```

```{r}
# OR, to save some typing 
fv('^b', data1)
```

Searching for concepts (aka the labels)

```{r}
# ?find_lab
# Find all the variable labels that mention "income"
find_lab('job', 
         case.sensitive = FALSE)
```

```{r}
# OR, to save some typing 
fl('depression')
```

You can also select variable you computed or renamed (more in the docs)

```{r}
# Select some variables with common root
select_var('depre', times=c(1,3)) # returns, for example c('depre_1','depre_3')

# OR, to save some typing 
sel('depre')
```

## 3. Clean dataset

#### Create a unique identifier

> [Attention]{.underline} what looks like the id variable is **not** a unique identifier!
> You need to make this by combining the `cidb` variable with `qlet`. Can do just that
> with **`make_idc()`**

-   same as before, if your data is called "data" that saves some typing

```{r}
?make_idc
# Create unique identifier
cbind(idc = make_idc(mom.id='cidb1234'), data)

# If you don't care about the position
# data$IDC <- make_idc(mom.id='cidb1234') 
```

#### Select siblings

Often you want to exclude related individuals (i.e. one of the two) from the analyses
**`rm_siblings()`** helps with that

-   it looks for a unique identifier (called `"idc"` by default) of the type `make_idc()`
    outputs. If he doesn't find it, calls `make_idc()` internally (using the `mom.id` and
    `parity` optional variable names)
-   `method` is the most useful argument I think, it allows to control who to exclude.
    Default is **"random"** = one sibling is randomly selected for each pair. **"parity"**
    = the oldest sibling is retained in the dataset. **"missing"** = the sibling with the
    lowest number of missing values is retained. Missing rates are computed over the
    entire data dataset or only over the columns specified in `column_selection`.
-   as usual "data" is the default name of the dataframe

```{r}
# ?rm_siblings
# Remove one of the sibling couple randomly
data_nosibl <- rm_siblings(
  # idc = "idc",
  method = "random",
  # column_selection = "all",
  # seed = 310896,
  # mom.id = "cidb1234",
  # parity = "qlet",
  # data = NULL
)
```

#### Remove outliers

```{r}
# ?rm_outliers

# Remove outliers 
data$horses <- rm_outliers('unicorns', cutoff=3.5)
```

## 4. Scoring, recoding and other fun stuff

-   `dichotomize()`

-   `dichotomize_cont()`

-   `ccei_score()`

-   `epds_score()`
